Investigating the Impact of Macroeconomic Indicators on the FTSE 250 Index Using Statistical and Machine Learning Methods

### 2024-08-24 tex/paper.tex
literature review revised

The remainder of this paper will be organised as follows. 
Section 2 provides a brief review of the literature in this problem space, and explains the rationale behind the selection of macroeconomic variables.
Section 3 presents the methodology of the paper. Section 4 discusses the results and Section 5 concludes
the study.

@@

The Fisher hypothesis states that nominal interest rates adjust to expected 
inflation as firms correctly anticipate inflation 
and adjust prices, implying that markets are efficient and that inflation has no impact on the stock market. This was tested this in the US stock market by \citeboth{jaffe1976}, and
\citeboth{bodie1976} who both found a negative relationship between inflation and stock returns, providing evidence against the Fisher hypothesis.
To account for this,
\citeboth{mogdiliani1979} proposed the Inflation Illusion Hypothesis, 
arguing that investors overestimate the discount rate in valuations, leading to mispricing. 
\citeboth{gultekin1983} and \citeboth{firth1979}, however, found that the relationship between nominal stock returns and inflation in the UK is positive, and that real returns on UK stocks have remained stable even as inflation varied.
Later, \citeboth{hasan2008} provided evidence against this, challenging the Fisher hypothesis and establishing a 
significant causal relationship between the stock market

@@

Interest rates, being the cost of borrowing, directly impact firms by increasing financing costs, reducing profitability, and lowering growth prospects. For consumers, higher rates raise borrowing costs, leading to reduced spending and demand, which further lowers corporate profitability \citep{boe199section}. 
The empirical research supports this theory. \citeboth{alam2009} found a significant inverse relationship between interest rates and stock prices across multiple countries from 1988 to 2003.
\citeboth{talla2013} observed a negative relationship between interest rates and stock prices on the Stockholm Stock Exchange.
\citeboth{demir2019} reported similar findings for the Turkish Stock Exchange. In the case of the UK, \citeboth{neifar2023} used a Non-Linear ARDL model to find a significantly negative long-run relationship between interest rates and the FTSE 100 Index. 
In addition, the flight-to-quality phenomenon that higher interest rates also encourage a shift from stocks to fixed-income 
assets such as bonds. \citeboth{asgharian2016} provided evidence for this in developed stock markets, including the UK. This suggests interest rates 
have a negative impact the stock market.

@@

Classical economic theory explains the impact exchange rates have on stock prices through the portfolio balance model. The portfolio balance model indicates that expected currency depreciations lead investors to shift from domestic to foreign assets, decreasing demand for domestic stocks and lowering their prices. 
\citeboth{branson1977} established a bidirectional relation between the exchange rate and stock prices. This was confirmed
in the case of the Turkish Stock Exchange by \citeboth{aydemir2009}. 
Moreover, stock markets often contain multinational firms involved in international transactions. Their profits are strongly influenced by the exchange rate, and consequently so is the stock market price \citep{Wong2018}.
In support of this, \citeboth{khan2018} found that the exchange rate showed a positive relationship with the Karachi Stock Exchange over the long-run, and a negative impact in the short-run. 
In the case of the UK, \citeboth{wong2022} also demonstrated a positive long-run impact by the exchange rate on the stock market.

@@

One of the foundational studies on the relationship between the money supply and 
stock market prices was \citeboth{palmer1970}, who analysed the impact of M1 money supply on the S$\&$P 425 Industrial Stock Index. 
Palmer posited that trends in the nation's money stock could lead the 
private sector to adjust wealth portfolios, resulting in 
predictable movements in stock prices. \citeboth{homa1971} further argued that 
stock prices are significantly impacted by the risk-free interest rate, 
which is a function of money supply. This relationship was disputed by \citeboth{pesando1974}, 
who critiqued the theoretical and empirical methodology used in their study. In addition, \citeboth{sorensen1982} and 
\citeboth{bernanke2005} argue that only the unanticipated component of money supply changes 
affects the stock market price. Later research, however, such as \citeboth{bahloul2017}, 
\citeboth{synek2024}, and \citeboth{pícha2017} has found that the price of developed stock markets, such as the US and UK's, is in fact significantly influenced by the money supply. 
The nature of this relationship is disputed and varies across countries. \citeboth{homa1971} established a positive relationship between the 
stock market and the money supply. This was disputed by \citeboth{sellin2001}, who presents a negative theoretical relationship betwen the stock market and money supply. \citeboth{olawale2014} empirically found a significant positive impact by the money supply on stock returns in the US, but a negative one in the case of the UK.

@@

>>>>

The remainder of this paper will be organised as follows. 
Section \ref{sec:lit} provides a brief review of the literature in this problem space, and explains the rationale behind the selection of macroeconomic variables.
Section 3 presents the methodology of the paper. Section 4 discusses the results and Section 5 concludes
the study.

@@

The impact of inflation on stock market prices has been a subject of debate. The Fisher hypothesis states that nominal interest rates adjust to expected 
inflation as firms correctly anticipate inflation 
and adjust prices, implying that markets are efficient and that inflation has no impact on the stock market. This was tested this in the US stock market by \citeboth{jaffe1976}, and
\citeboth{bodie1976} who both found a negative relationship between inflation and stock returns, providing evidence against the Fisher hypothesis.
To account for this,
\citeboth{mogdiliani1979} proposed the Inflation Illusion Hypothesis, 
arguing that investors overestimate the discount rate in valuations, leading to mispricing. 
\citeboth{gultekin1983} and \citeboth{firth1979}, however, found that the relationship between nominal stock returns and inflation in the UK is positive, and that real returns on UK stocks have remained stable even as inflation varied.
Later, \citeboth{hasan2008} provided evidence against this, challenging the Fisher hypothesis and establishing a 
significant causal relationship between the stock market

@@

The interest rate has been confirmed to have a definite effect on the 
stock market price by the findings of \citeboth{ChenRollRoss1986} and later research. 
Interest rates, being the cost of borrowing, directly impact firms by increasing financing costs, reducing profitability, and lowering growth prospects. For consumers, higher rates raise borrowing costs, leading to reduced spending and demand, which further lowers corporate profitability \citep{boe199section}. 
The empirical research supports this theory. \citeboth{alam2009} found a significant inverse relationship between interest rates and stock prices across multiple countries from 1988 to 2003.
\citeboth{talla2013} observed a negative relationship between interest rates and stock prices on the Stockholm Stock Exchange.
\citeboth{demir2019} reported similar findings for the Turkish Stock Exchange. In the case of the UK, \citeboth{neifar2023} used a Non-Linear ARDL model to find a significantly negative long-run relationship between interest rates and the FTSE 100 Index. 
In addition, the flight-to-quality phenomenon that higher interest rates also encourage a shift from stocks to fixed-income 
assets such as bonds. \citeboth{asgharian2016} provided evidence for this in developed stock markets, including the UK. This suggests interest rates 
have a negative impact the stock market.

@@

Economic theory posits that the exchange rate, through the portfolio 
balance model and flow model, has a definite impact on stock prices. The portfolio balance model indicates that expected currency depreciations lead investors to shift from domestic to foreign assets, decreasing demand for domestic stocks and lowering their prices (\citeboth{branson1977}). 
In addition, variations in the exchange rate affect future cash flows, which in turn affect the present value of a stock and thus stock prices (\citeboth{dornbusch1980}).
This was confirmed
in the case of the US and Turkish Stock Exchanges by \citeboth{aggarwal1981} and \citeboth{aydemir2009}, respectively. 
Moreover, stock markets often contain multinational firms involved in international transactions. Their profits are strongly influenced by the exchange rate, and consequently so is the stock market price \citep{Wong2018}.
In support of this, \citeboth{khan2018} found that the exchange rate showed a positive relationship with the Karachi Stock Exchange over the long-run, and a negative impact in the short-run. 
In the case of the UK, \citeboth{wong2022} also demonstrated a positive long-run impact by the exchange rate on the stock market.

@@

The nature of the impact of the money supply on the stock market price is also 
in debate. \citeboth{palmer1970} posited that trends in the nation's money stock could lead the 
private sector to adjust wealth portfolios, resulting in 
predictable movements in stock prices. \citeboth{homa1971} further argued that 
stock prices are significantly impacted by the risk-free interest rate, 
which is a function of money supply. This relationship was disputed by \citeboth{pesando1974}, 
who critiqued the theoretical and empirical methodology used in their study. In addition, \citeboth{sorensen1982} and 
\citeboth{bernanke2005} argue that only the unanticipated component of money supply changes 
affects the stock market price. Later research, however, such as \citeboth{bahloul2017}, 
\citeboth{synek2024}, and \citeboth{pícha2017} has found that the price of developed stock markets, such as the US and UK's, is in fact significantly influenced by the money supply. 
The nature of this relationship is disputed and varies across countries. \citeboth{homa1971} established a positive relationship between the 
stock market and the money supply. This was disputed by \citeboth{sellin2001}, who presents a negative theoretical relationship betwen the stock market and money supply. \citeboth{olawale2014} empirically found a significant positive impact by the money supply on stock returns in the US, but a negative one in the case of the UK.

@@

Initially, the FTSE 250's composition was broadly distributed across 
various industrial and service sectors. Roughly $47\%$ of its constituents 
were involved in the industrial and utilities sectors, reflecting the UK's 
slower transition from an industrial-based to a service-based economy. 
Over time, the FTSE250 has evolved to align more closely with the 
structural transformation of the UK economy. 
By 2022, 73.39$\%$ of the index was comprised of service firms, 
with a significant 43.37$\%$ of these being financial services companies (\citeboth{ftse250history}) .

As a result, the relevance of the selected macroeconomic indicators for the FTSE250 
has increased. The M3 money supply and interest rates significantly 
impact the ability of financial firms 
to provide loans, facilitate investments, and manage assets, thereby affecting their profitability. 
Financial services firms also play a crucial role in supporting the 
operations of firms in all sectors of the economy, meaning 
changes in the interest rate and money supply affects their profitability too. 
The CPI is also of critical importance across all sectors, 
as inflation, measured by the CPI, erodes the purchasing power of money. 
This erosion diminishes consumer spending and negatively affects the 
profitability of service-oriented companies that depend on consumer 
expenditure. The exchange rate is especially relevant for transnational 
firms within the FTSE250. The profits of these companies, which make up a substantial 
portion of the index, are directly impacted by fluctuations in exchange 
rates that affect the value of their earnings.
